---
title: "Response to reviewer comments"
authors: Nicholas Tierney, Karthik Ram
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r libs, include = FALSE}
library(dplyr)
library(googlesheets4)
library(knitr)
library(kableExtra)
```


```{r read-sheet, include = FALSE}
our_table_id <- gs4_get(
  "14-qpuM3DaeaeWt4D6E7dLuvkT3L4QTe2gDr3ObxveOI"
  )

our_table <- read_sheet(ss = our_table_id,
                        .name_repair = janitor::make_clean_names) %>% 
  select(ID = id, 
         comment = reviewer_comment, 
         response) %>% 
  tidyr::drop_na()
```

Dear Dr. Callaghan,

Thank you for accepting our commentary article, "Common sense approaches to sharing tabular data alongside publication", for publication in Patterns.

We thank the reviewers for time and detailed comments. We have addressed the final comments from the reviewers, our response to individual reviewer comments are presented in the table on page 2 of this letter. 

Regards,

Nicholas Tierney & Karthik Ram

\newpage

```{r show-table}
our_table %>% 
  kbl(booktabs = TRUE,
      longtable = TRUE) %>% 
  kable_styling(latex_options = c("striped",
                                  "hold_position", 
                                  "repeat_header"),
                full_width = TRUE) %>% 
  column_spec(1) %>% 
  column_spec(2, width = "20em") %>% 
  column_spec(3, width = "20em")
```

